---
title: "GeoNonStat"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
author: "Sébastien Coube-Sisqueille & Elise Maigné"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Your Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
options(warn=-1)
```

# Introduction

The aim of GeoNonStat is non-stationary geostatistical modeling. 

The data is decomposed as 
$$ y(s) = x(s)^T\beta + w(s) + \epsilon(s),$$

 - $s$ being a spatial coordinate in 2D,
 - $y(s)$ being a univariate continuous response variable at the coordinate $s$,
 - $x(s)$ being a vector of explanatory variables at the coordinate $s$,
 - $\beta$ being a vector of regression coefficient,
 - $w(s)$ being an uni-variate spatial effect term at the coordinate $s$ who captures the spatial auto-correlation of the data,
 - $\epsilon(s)$ being an uni-variate noise term at the coordinate $s$.
 
Measurements are done in several coordinates $S = (s_1, \ldots, s_n)$, and replicates are possible. 

Both $w(s)$ and $\epsilon(s)$ are nonstationary: their behavior may change following the coordinate $s$ and some explanatory variables. 

### Regression model for the explanatory variables

It works like an usual linear regression model, in lm() for example. 
The estimates or confidence intervals may change with respect to an usual linear model due to the spatial effects. 

### Nonstationary Gaussian Process model for the spatial effect

The spatial effect is specified through a *Gaussian Process* who guarantees its spatial coherence. 
$$w(S) \sim \mathcal{N}\left(0,~\sigma^2~M(S, \theta(S), \nu)\right)$$

While the mean of $w(S)$ is null to avoid identifiability problems with 
the regression model, its covariance matrix is decomposed into two components: 

- $\sigma^2$ is the marginal variance parameter, that is $Var(w(s_1)) = \ldots = Var(w(s_n)) = \sigma^2$.
- $M(S, \theta(S),\nu)$ is the correlation matrix of $w(S)$. It depends on three things: 
  - Spatial coordinates $S$. Roughly, the closer $s_i$ and $s_j$ are, the more correlated $w(s_i)$ and $w(s_j)$.
  - Spatial correlation range parameters $\theta(S)$, who control the local correlations structure of $w(S)$. 
  Those parameters are the reason why $w(S)$ is nonstationary (or not). 
  - A smoothness parameter $\nu$ who can take value $0.5$ or $1.5$. If $\nu=0.5$, the spatial field will be a bit rougher. 

#### Range and smoothness

Imagine $w(S)$ as a landscape of hills. The width of the hills is controlled by the range $\theta$, and their roughness is controlled by the smoothness $\nu$. The smoothness is an important parameter, but it is very difficult to identify with the range, so the smoothness is set by the user to $0.5$ or $1.5$. 

`````{r, echo=FALSE, message=FALSE, results='hide'}
library(GeoNonStat)
observed_locs = as.matrix(expand.grid(seq(0, 1, .005), seq(0, 1, .005)))
vecchia_approx = createVecchia(observed_locs, m = 6, round_locs = 0)
PP_range = createPP(vecchia_approx, matern_range = .5, plot = F)
field_log_var=  0
range_log_scale = c(0, 0)
noise_intercept = -2

par(mfrow  =c(2,2))
for(nu in c(.5, 1.5)){
  gns_simulator = createGnsSimulator(
    vecchia_approx = vecchia_approx,
    range_PP = PP_range, 
    anisotropic = T, matern_smoothness = nu
  )
  rangevalues = c(-4, -2)
  for(beta0 in rangevalues){
    set.seed(1)
    coeff_list = createGnsSimulatorParameters(
      gns_simulator, range_PP_log_var =range_log_scale, 
      noise_intercept = noise_intercept, noise_PP_log_var = noise_PP_log_var, 
      verbose = F, field_log_var = field_log_var)
    coeff_list$range_X_coeff[1,1] = beta0
    coeff_list$range_PP_coeff[] = 0
    fake_data = simulateGnsData(gns_simulator = gns_simulator, gns_params = coeff_list)
    plotPointillistPainting(
      vecchia_approx$locs, fake_data$latent_field, cex= .5, 
      pch= 15, main = paste(c("small", "big")[match(beta0, rangevalues)], "range,", c("small", "big")[match(nu, c(.5, 1.5))], "smoothness"), legend = F, override_layout = F)
    print(nu)
    print(beta0)
  }
}
`````  
  
#### Spatial variation of the spatial effect's range

Now imagine that our landscape behaves differently following the region and/or some explanatory variables. 
The range parameters of the spatial effect are decomposed as: 
 $$\theta(s) = x_\theta(s)^T\beta_\theta +  w_\theta(s)$$
 
 - $s$ being a spatial coordinate in 2D,
 - $x_\theta(s)$ being a vector of explanatory variables at the coordinate $s$,
 - $\beta_\theta$ being a vector of regression coefficient,
 - $w(s)$ being an spatial effect term at the coordinate $s$ who enforces some spatial coherence on $\theta$. 
 
This differs from the decomposition of the data by the fact that there is no noise. 

The spatial coherence on $w_\theta$ is enforced by a Gaussian Process, with distribution: 
$$w_\theta(S) \sim \mathcal{N}\left(0,~\sigma^2_\theta~M(S,\alpha,\nu = 1.5)\right)$$

- $\sigma^2_\theta$ is the marginal variance parameter.
- $M(S,\alpha_\theta,\nu = 1.5)$ is the correlation matrix of $w(S)$. Now, 
  - The range $\alpha$ is constant and chosen by the modeler
  - The smoothness $\nu$ is set to $1.5$. 

So one parameter ($\alpha_\theta$) is user-chosen, and one ($\sigma_\theta$) is estimated by the model. 

For the sake of identifiability, the spatial coherence of $\theta(S)$ must be much higher than the spatial coherence of $w(S)$. There are two practical consequences to that: 

- $\alpha_\theta$ must be big,
- $X_\theta$ must be really smooth and spatially coherent. 

When $X_\theta$ is just an intercept, and $\sigma^2_\theta$ is constrained to $0$, then $\theta(S)$ is constant and the model boils down to a stationary model. 
`````{r, echo=FALSE, message=FALSE, results='hide'}
par(mfrow  =c(1,2))
  for(nonstationary in c(FALSE, TRUE)){
    set.seed(1)
    coeff_list = createGnsSimulatorParameters(
      gns_simulator, range_PP_log_var =range_log_scale, 
      noise_intercept = noise_intercept, noise_PP_log_var = noise_PP_log_var, 
      verbose = F, field_log_var = field_log_var)
    coeff_list$range_X_coeff[1,1] = -4
    coeff_list$range_PP_coeff[,-1] = 0
    coeff_list$range_PP_coeff[,1] = nonstationary * rnorm(length(coeff_list$range_PP_coeff[,1]))
    fake_data = simulateGnsData(gns_simulator = gns_simulator, gns_params = coeff_list)
    plotPointillistPainting(
      vecchia_approx$locs, fake_data$latent_field, cex= .5, 
      pch= 15, main = paste(c("stationary", "nonstationary")[1+nonstationary]), legend = F, override_layout = F)
    print(nu)
    print(beta0)
  }
````` 


#### Local isotropy or anisotropy of the spatial effect

The spatial effect can be either locally isotropic or anisotropic. 
When the correlation of the spatial effect does not change with spatial orientation, the field is isotropic. 
If not so, the field is anisotropic. The model can be anisotropic and nonstationary, with both the correlation range and direction evolving across space. 

`````{r, echo=FALSE, message=FALSE, results='hide'}
par(mfrow  =c(1,3))
for(aniso in c(FALSE, TRUE)){
  for(nonstationary in c(FALSE, TRUE)[seq(1+ aniso)]){
    set.seed(1)
    coeff_list = createGnsSimulatorParameters(
      gns_simulator, range_PP_log_var =range_log_scale, 
      noise_intercept = noise_intercept, noise_PP_log_var = noise_PP_log_var, 
      verbose = F, field_log_var = field_log_var)
    coeff_list$range_X_coeff[1,1] = -4
    coeff_list$range_X_coeff[1,2] = .3*aniso
    coeff_list$range_X_coeff[1,3] = -.6*aniso
    coeff_list$range_PP_coeff[] = rnorm(length(coeff_list$range_PP_coeff[]))*nonstationary
    fake_data = simulateGnsData(gns_simulator = gns_simulator, gns_params = coeff_list)
    plotPointillistPainting(
      vecchia_approx$locs, fake_data$latent_field, cex= .5, 
      pch= 15, main = paste(c("isotropic", "anisotropic")[1+aniso], "and", c("stationary", "nonstationary")[1+nonstationary]), legend = F, override_layout = F)
    print(nu)
    print(beta0)
  }
}
`````  
Notes: 

  - The model choice changes the dimension of $\theta(s)$:
    - The isotropic model is parametrized using scalars, so $\theta(s)$ is a vector with length 1.
    - The anisotropic model is parametrized using ellipses, so $\theta(s)$ is a vector with length 3.
  - When the model is nonstationary, is is by definition not isotropic. However, since $X\theta$ and $w_\theta(S)$
  are supposed to be much more coherent than $w(S)$, the we can consider than the field is locally isotropic. 

#### Takeaway

- The latent field can be either (locally) isotropic or anisotropic.
- The latent field can be either stationary or nonstationary.
- If the latent field is nonstationary: 
    - The explanatory variables $x_\theta(s)$ explaining the spatial range of the spatial effect must be highly coherent spatially.
    - The spatial effects $w_\theta(s)$ explaining the spatial range of the spatial effect must be highly coherent spatially too.


## Nonstationary model for the noise

## Model architecture and catalogue of the parameters



#### Parameters, set by the user
#### Hyperparameters, set by the user